---
title: "Using the MUTT data API"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using the MUTT data API}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE, purl = FALSE)
```

MUTT has one exported function, `mutt()`. The package bundles parser code while study data are resolved from a local Git LFS checkout or a checksum-verified user cache.

```{r, purl = FALSE}
library(mutt)

x <- mutt(
  studies = "2021_vandeputte_naturecommunications_flow_timeseries",
  base_directory = "/path/to/mutt"
)
```

The returned object remains a named list of study objects. Audit information is additive and does not replace study entries.

```{r, purl = FALSE}
x[[1]]$counts
x[[1]]$scale
attr(x, "audit")
attr(x, "validation")
attr(x, "provenance")
```

## Resolution and execution

1. Validate arguments and registered study IDs.
2. Resolve each study from `base_directory`, `MUTT_DATA_DIR`, `./studies`, or the user cache.
3. Download and verify a published study bundle only when no local copy exists.
4. Load the parser bundled under `inst/parsers`.
5. Run the unchanged parser in a scoped data-root context.
6. Validate the returned object.
7. Optionally run or reuse PICRUSt2 and FAPROTAX.
8. Return successful studies with structured status, validation, and provenance attributes.

The scoped parser context is a compatibility boundary for historical parsers that use study-relative paths. It restores the caller's working directory after success or failure.

## Functional inference

```{r, purl = FALSE}
x <- mutt(
  studies = "2022_cvandevelde_ismecommunications_culturedflowhumanfecal",
  functional = TRUE
)
```

`functional = TRUE` reuses matching caches. `functional = "REBUILD"` recomputes eligible branches. Neither mode rarefies counts, adds pseudocounts, closes compositions, or log-transforms the input counts.

PICRUSt2 is used only when integer ASV counts, aligned sequences, and aligned prokaryotic taxonomy are available. FAPROTAX uses a matching classified taxonomic abundance table: integer counts are preferred, and proportions are used only when counts are unavailable. Missing external tools produce structured skipped records rather than changing parsed study data.

Stratified PICRUSt2 output remains file-backed:

```{r, purl = FALSE}
branch <- x[[1]][["function"]]$picrust2[[1]]
head(as.data.frame(branch, type = "ko", n_max = 100))
```

## Failure interpretation

A parser error is not converted into a successful empty dataset. MUTT continues with other requested studies, leaves the failed study entry `NULL`, emits one summary warning, and records the error in `attr(x, "audit")`.

A functional-engine failure does not invalidate an otherwise valid parsed study. It is recorded in the study's functional manifest and audit warnings while counts, scale, metadata, and taxonomy remain available.
